Galaxy Zoo: the dependence of the star formation-stellar mass relation on spiral disk morphology

### 2014-12-03 paper/mainsequence_morph.tex
Final typo check

\begin{abstract}
We measure the mass-star formation relation in star-forming disk galaxies at $z<0.05$, using Galaxy~Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their average pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only boost the SFR above the main relation by $\sim0.3$~dex; this is significantly smaller than the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the various quenching mechanisms or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed, but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

As a star-forming galaxies may exhibit a wide range of physical appearances in optical images, the natural question can be asked of whether the specifics of this physical appearance and its underlying dynamical processes have any effect on this homeostasis and therefore the galaxy's location relative to the SFMS. If the details of a galaxy's physical appearance are correlated with position relative to the main sequence, then the dynamical processes that give rise to them -- such as bar formation and spiral arm pitch angle -- are a fundamental aspect of the galaxy's regulatory mechanism. If, on the other hand, these features are not correlated, then there are two options: either galaxy substructure is simply not relevant to the overall $M_\star$-SFR relationship, or the regulatory mechanism overcomes the local effect of substructure in all circumstances. This ultimately relates to the physical processes that control the overall strength of the regulator in each galaxy.

The fact that star-forming galaxies live on the SFMS is one of the key observations that has been driving the development of new descriptions of how galaxies evolve \citep[eg, ][]{sch07b}. \citet{pen10a,pen12} argue that galaxies grow in stellar mass during their life as star-forming galaxies on the main sequence before having their star formation quenched either by an external mechanism (`environment quenching') or by an internal mechanism (`mass quenching').  However, \citet{del12} point out that, because of a `history bias,' galaxies of identical stellar mass may have different environmental histories that are difficult to disentangle, thus making the mass versus environment debate inherently ill-posed \citep[see also ][]{van08b}. In addition, Galaxy~Zoo has found that the environmental dependence of galaxy star formation and that of morphology are not equivalent, mainly because of the abundance of quenched spiral galaxies, a large fraction of which are satellite galaxies \citep{ski09,bam09}. In any case, life on the main sequence appears to be governed by the action of the regulator to balance gas in- and outflows \citep{lil13}, making it a central process in galaxy evolution.

@@

Spiral galaxies are identified according to the following thresholds in the spectroscopic GZ2 sample, where $p$ is the debiased vote fraction and $N$ the weighted number of total votes: $p_\textrm{features/disk} > 0.430$, $p_\textrm{not~edgeon} > 0.715$, $p_\textrm{spiral}>0.619$, and $N_\textrm{spiral}>20$. We identify sub-classes of spiral structure by weighting each galaxy according to the fraction of votes received in each morphological category.

@@

@@

As a control sample, we also analyze the fits to the underlying star-forming main sequence relation for star-forming disks as measured in a sample of local SDSS galaxies (Figures~\ref{fig-number}-\ref{fig-bar},\ref{fig-mergers}). As previously demonstrated with SDSS data \citep[e.g.,][]{bri04}, there is a tight correlation between $M_\star$ and SFR, with galaxies in the process of quenching lowering their SFR and falling below the trend. The relationship extends over at least 4~orders of magnitude in both $M_\star$ and SFR. Fits to the SFMS for all subsamples in this paper are listed in Table~\ref{tbl-fits}.

@@

The fit to the star-forming main sequence for all six sub-samples of spiral galaxies split by arm multiplicity tightly follows that of the total star-forming disk population. Both the slopes and offsets of each linear fit are consistent within the formal fitting errors (Table~\ref{tbl-fits}). There are too few galaxies with reliable spiral morphologies at lower masses to do a reliable fit; however, the same relation is reproduced for all multiplicities at $10^8<M_\star/M_\odot<10^{11}$.

One-armed spiral galaxies present an interesting case, with the fit to the weighted population lying slightly \emph{above} that of all star-forming spirals. This is consistent with the work of \citet{cas13}, who showed that one-armed spirals in GZ2 are robust indicators of close interactions at projected distances of $r_p < 50~h^{-1}$~kpc. The underlying reason is that many ``one-armed spirals'' are in fact caused by bridges or tidal tails from interactions with a nearby companion instead of secular processes. We discuss the likely role of merging/interacting galaxies in Section~\ref{sec-discussion} (see Figure~\ref{fig-mergers}).

@@

We have repeated the analysis above for the sub-sample of disk galaxies for which the spiral multiplicity is determined with high confidence ($p_\textrm{arms~number} > 0.8$) by GZ2, thus eliminating ``intermediate'' galaxies for which the morphology is uncertain. These galaxies ($N=14,148$) are dominated by two-armed spirals, which are the only multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\odot$ are detected. Our results for the SFMS are qualitatively the same as when using the weighting scheme, although we note that there are too few examples ($N<10$) of either three- or four-armed spirals for a reliable fit. The offset of the one-armed spirals above the SFMS is also significantly more pronounced when using high-confidence morphologies.

The pitch angle of the spiral arms also has no significant change on the star-forming main sequence relation (Figure~\ref{fig-winding}). We separate galaxies by their relative pitch angles (defined as ``tight'', ``medium'', and ``loose''); the pitch angle is typically used as one of the primary parameters for separating galaxies along the Hubble tuning fork. \citet{wil13} show, however, that pitch angle only weakly correlates with Hubble type from expert visual classifications, and that the bulge-to-disk ratio is a more important driver. There is no significant shift with respect to the SFMS as a function of pitch angle for spiral galaxies. We note that the small increase above the main sequence for loosely-wound galaxies is also consistent with \citet{cas13}, who show that this morphology also correlates with close pairs and interactions.

\begin{figure*}
\includegraphics[angle=0,width=7.0in]{figures/ms_arms_winding.pdf}
\caption{Total star formation rate as a function of stellar mass; grayscale colors are the same as in Figure~\ref{fig-number}. From left to right: red, green, and blue points are spiral galaxies with ``tight'', ``medium'', and ``loose'' winding spiral arms as identified by GZ2 morphology flags. Dotted lines show the weighted least-squares linear fit as split by pitch angle; the solid line is the fit to all star-forming galaxies. The positive offset in SFR for loosely-wound spiral arms is interpreted as contamination by merging pairs of galaxies (Section~\ref{sec-discussion}).
\label{fig-winding}}
\end{figure*}

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compared the sample standard deviation ($\sigma_{SFR}$) as a function of mass of the star-forming galaxy population to its various subsamples. The value of $\sigma_{SFR}$ monotonically decreases with increasing mass over the range $6.0<\log(M/M_\odot)<11.5$. For all populations examined in this paper, though, the widths of their distributions are consistent with the broader population (Figure~\ref{fig-sigma}).

@@

We have examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in the results.

@@

As a comparison to the kpc-scale structures discussed above, we analyze the impact of the most significant forcing event to a galaxy system known -- a major galaxy merger (Figure~\ref{fig-mergers}). In these systems, which are in various stages of coalescence, star formation rates are increased by only an average of 0.29~dex (less than a factor of two). \citet{dar10} showed that in the local ($z<0.1$) Universe, galaxies with intense bursts of star formation are limited to only the spiral (disc) galaxies. This increase in star formation for mergers does show a strong evolution in redshift out to at least $1.5<z<2.5$, likely due to the higher gas fractions involved \citep{dad10,rod11}. The projected separation between galaxies in which this occurs, based on both the observed merger fraction and sSFR is $\sim0.1$~Mpc$/h$ \citep{ski09}. Our measurements are consistent with observations of galaxies at $z\simeq2$ \citep{kav13b}, which support a merger-driven increase of only a factor of $\sim2$ in sSFR.

>>>>

\begin{abstract}
We measure the mass-star formation relation in star-forming disk galaxies at $z<0.10$, using Galaxy~Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their average pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only boost the SFR above the main relation by $\sim0.3$~dex; this is significantly smaller than the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the various quenching mechanisms or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed, but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

As star-forming galaxies may exhibit a wide range of physical appearances in optical images, the natural question can be asked whether the specifics of this morphology and its underlying dynamical processes have any effect on this homeostasis and therefore the galaxy's location relative to the SFMS. If the details of a galaxy's physical appearance are correlated with position relative to the main sequence, then the dynamical processes that give rise to them -- such as bar formation and spiral arm pitch angle -- are a fundamental aspect of the galaxy's regulatory mechanism. If, on the other hand, these features are not correlated, then there are two options: either galaxy substructure is simply not relevant to the overall $M_\star$-SFR relationship, or the regulatory mechanism overcomes the local effect of substructure in all circumstances. This ultimately relates to the physical processes that control the overall strength of the regulator in each galaxy.

The fact that star-forming galaxies live on the SFMS is one of the key observations that has been driving the development of new descriptions of how galaxies evolve \citep[eg, ][]{sch07b}. \citet{pen10a,pen12} argue that galaxies grow in stellar mass during their life as star-forming galaxies on the main sequence before having their star formation quenched either by an external mechanism (`environment quenching') or by an internal mechanism (`mass quenching').  However, \citet{del12} point out that, because of a `history bias,' galaxies of identical stellar mass may have different environmental histories that are difficult to disentangle, thus making the mass versus environment debate inherently ill-posed \citep[see also][]{van08b}. In addition, Galaxy~Zoo data has shown that the environmental dependence of galaxy star formation and that of morphology are not equivalent, mainly because of the abundance of quenched spiral galaxies, a large fraction of which are satellite galaxies \citep{ski09,bam09}. In any case, life on the main sequence appears to be governed by the action of the regulator to balance gas in- and outflows \citep{lil13}, making it a central process in galaxy evolution.

@@

@@

The spiral nature of the star-forming disk galaxies are identified according to the following thresholds in the GZ2 spectroscopic sample, where $p$ is the debiased vote fraction and $N$ the weighted number of total votes: $p_\textrm{features/disk} > 0.430$, $p_\textrm{not~edgeon} > 0.715$, $p_\textrm{spiral}>0.619$, and $N_\textrm{spiral}>20$. We identify sub-classes of spiral structure by weighting each galaxy according to the fraction of votes received in each morphological category.

@@

As a control sample, we also analyze the fits to the underlying star-forming main sequence relation for star-forming disks as measured in a sample of local SDSS galaxies (Figures~\ref{fig-number}-\ref{fig-bar}, \ref{fig-mergers}). As previously demonstrated with SDSS data \citep[e.g.,][]{bri04}, there is a tight correlation between $M_\star$ and SFR, with galaxies in the process of quenching lowering their SFR and falling below the trend. The relationship extends over at least 4~orders of magnitude in both $M_\star$ and SFR. Fits to the SFMS for all subsamples in this paper are listed in Table~\ref{tbl-fits}.

@@

The fit to the star-forming main sequence for all six sub-samples of spiral galaxies split by arm multiplicity tightly follows that of the total star-forming disk population. Both the slopes and offsets of each linear fit are consistent within the formal fitting errors (Table~\ref{tbl-fits}). There are too few individual galaxies with confirmed spiral morphologies at lower masses to do a reliable fit; however, the same relation is reproduced for all arm multiplicities at $10^8<M_\star/M_\odot<10^{11}$.

One-armed spiral galaxies present an interesting case, with the fit to the weighted population lying slightly \emph{above} that of all star-forming spirals. This is consistent with the work of \citet{cas13}, who showed that one-armed spirals in GZ2 are robust indicators of close interactions at projected distances of $r_p < 50~h^{-1}$~kpc. The underlying reason is that many ``one-armed spirals'' are in fact caused by bridges or tidal tails from interactions with a nearby companion instead of secular processes. We discuss the likely role of merging/interacting galaxies in Section~\ref{sec-discussion} (also see Figure~\ref{fig-mergers}).

@@

We have repeated the analysis above for the sub-sample of disk galaxies for which the spiral multiplicity is determined with high confidence ($p_\textrm{arms~number} > 0.8$) by GZ2, thus eliminating ``intermediate'' galaxies for which the morphology is uncertain. These galaxies ($N=14,148$) are dominated by two-armed spirals, which are the only spiral multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\odot$ are detected. The results for the SFMS are qualitatively the same as when using the weighting scheme, although we note that there are too few examples ($N<10$) of either three- or four-armed spirals for a reliable fit. The offset of the one-armed spirals above the SFMS is also significantly more pronounced when using high-confidence morphologies.

The pitch angle of the spiral arms also has no significant change on the star-forming main sequence relation (Figure~\ref{fig-winding}). We separate galaxies by their relative pitch angles (defined as ``tight'', ``medium'', and ``loose''); the pitch angle is typically used as one of the primary parameters for separating galaxies along the Hubble tuning fork. \citet{wil13} show, however, that pitch angle only weakly correlates with Hubble type from expert visual classifications, and that the bulge-to-disk ratio is a more important driver. There is no significant shift with respect to the SFMS as a function of pitch angle for spiral galaxies. The small increase above the main sequence for loosely-wound galaxies is also consistent with \citet{cas13}, who show that this morphology also correlates with close pairs and interactions.

\begin{figure*}
\includegraphics[angle=0,width=7.0in]{figures/ms_arms_winding.pdf}
\caption{Total star formation rate as a function of stellar mass; grayscale colors are the same as in Figure~\ref{fig-number}. From left to right: red, green, and blue points are spiral galaxies with ``tight'', ``medium'', and ``loose'' winding spiral arms as identified by GZ2 morphology flags. Dotted lines show the weighted least-squares linear fit as split by pitch angle; the solid line is the fit to all star-forming galaxies. The slight positive offset in SFR for loosely-wound spiral arms is interpreted as contamination by merging pairs of galaxies (Section~\ref{sec-discussion}).
\label{fig-winding}}
\end{figure*}

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compare the sample standard deviation ($\sigma_{SFR}$) to the star-forming galaxy population over its various morphological subsamples. The value of $\sigma_{SFR}$ monotonically decreases with increasing mass over the range $6.0<\log(M/M_\odot)<11.5$. For all populations examined in this paper, though, the widths of their distributions are consistent with the broader population (Figure~\ref{fig-sigma}).

@@

We have examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in any of the results.

@@

As a comparison to the kpc-scale structures discussed above, we analyze the impact of the most significant forcing event to a galaxy system known -- a major galaxy merger (Figure~\ref{fig-mergers}). In these systems, which are in various stages of coalescence, star formation rates are increased by only an average of 0.29~dex (less than a factor of two). \citet{dar10} showed that at $z<0.1$, galaxies with intense bursts of star formation are limited to only the spiral (disc) galaxies. This increase in star formation for mergers does show a strong evolution in redshift out to at least $1.5<z<2.5$, likely due to the higher gas fractions involved \citep{dad10,rod11}. The projected separation between galaxies in which this occurs, based on both the observed merger fraction and sSFR is $\sim0.1$~Mpc$/h$ \citep{ski09}. Our measurements are consistent with observations of galaxies at $z\simeq2$ \citep{kav13b}, which support a merger-driven increase of only a factor of $\sim2$ in sSFR.
